Performance of time series, case-crossover, and cross-year models in estimating the effects of natural disasters on human health in the presence of delayed or displaced effects

### 2016-06-24 writing/JSSDraft/JSSDraft.Rmd
Add some references to Bibtex file
Work on paper drafts

---
author:
  - name: Sarah Koehler
    affiliation: Colorado State University
    address: >
      Lake Street
      Fort Collins, CO
    email: sakoehler7@gmail.com
  - name: G. Brooke Anderson
    affiliation: Colorado State University
    address: >
      Lake Street
      Fort Collins, CO
    email: brooke.anderson@colostate.edu
title:
  formatted: "\\pkg{eesim}: An R Package to Simulate Environmental Epidemiology Time Series for Simulation Studies and Power Analyses"
  # If you use tex in the formatted title, also supply version without
  plain:     "eesim: An R Package to Simulate Environmental Epidemiology Time Series for Simulation Studies and Power Analyses"
  # For running headers, if needed
  short:     "\\pkg{eesim}: Simulations for Environmental Epidemiology"
abstract: >
  The abstract of the article.
keywords:
  # at least one keyword must be supplied
  formatted: [R, environmental epidemiology, simulation studies, power analysis, air pollution]
  plain:     [R, environmental epidemiology, simulation studies, power analysis, air pollution]
preamble: >
  \usepackage{amsmath}
output:
  rticles::jss_article:
    citation_package: natbib
bibliography: mybibfile.bib
---

@@

* Not consistent across different studies
* Time consuming to create, often only need to customize certain parts

@@

Example citation [@Dirac1953888].

>>>>

---
author:
  - name: Sarah Koehler
    affiliation: Colorado State University
    address: >
      Lake Street
      Fort Collins, CO
    email: sakoehler7@gmail.com
  - name: G. Brooke Anderson
    affiliation: Colorado State University
    address: >
      Lake Street
      Fort Collins, CO
    email: brooke.anderson@colostate.edu
title:
  formatted: "\\pkg{eesim}: An R Package to Simulate Environmental Epidemiology Time Series for Simulation Studies and Power Analyses"
  # If you use tex in the formatted title, also supply version without
  plain:     "eesim: An R Package to Simulate Environmental Epidemiology Time Series for Simulation Studies and Power Analyses"
  # For running headers, if needed
  short:     "\\pkg{eesim}: Simulations for Environmental Epidemiology"
abstract: >
  Simulation studies are important in environmental epidemiology research on air pollution, temperature, and other exposures. For example, studies have simulated time series of environmental exposure and health outcome data to compare time series and case-crossover methods; explore models to test for mortality-displacement; investigate the impacts of measurement error and spatial misalignment of exposure and outcome measurements; and search for confounding by other time-varying covariates.

  Developing the code to simulate environmental time series for these types of studies can be time consuming, and simulation methods have varied across previous simulation studies, complicating the comparison of results across different studies. Here, we present an R package that automates simulation of environmental time series for such simulations, offering both sensible defaults for each stage of the simulation process and also the option to customize any of the simulation steps while using defaults for any steps that do not require customization. This packge can be used to aid simulation studies to explore environmental epidemiology models relating acute health outcomes and daily environmental exposures; it can also be used to estimate the power of studies when planning or proposing future research.
keywords:
  # at least one keyword must be supplied
  formatted: [R, environmental epidemiology, simulation studies, power analysis, air pollution]
  plain:     [R, environmental epidemiology, simulation studies, power analysis, air pollution]
preamble: >
  \usepackage{amsmath}
output:
  rticles::jss_article:
    citation_package: natbib
bibliography: mybibfile.bib
---

@@

For many environmental time series studies, including those investigating air pollution and temperature, both exposures and outcomes can have complex seasonal and longer-term patterns over the study period. For example, air pollution concentrations can change seasonally, both because of changes in emission patterns and meteorological patterns (for example, winter inversion layers can cause a build-up of particulate matter in the winter, while high temperatures in the summer can accelate ozone formation from precursors); air pollution concentrations can also change over years, especially with evolving Clean Air policies.

Exposures, outcomes, and their relationship with each other can all have complex patterns that could cause problems in epidemiological studies. For example, \citet{Armstrong2014} explain how autocorrelation in temperature data (e.g., days right after a very severe heat wave are likely to be hotter than days right after a milder heat wave) can introduce spurious results in a model testing for evidence of short-term mortality displacement following heat waves.

There are a few important advantages to having an open-source software package that can simulate this type of environmental time series for simulation studies. First, current methods of simulating environmental time series, to use to assess modeling choices, vary across different studies. While, ideally, results from a simulation study would be robust to different choices in how to simulate the data used in the study, this may not always be the case; if so, results across studies using different methods of simulating data would not be comparable. This package offers a tool that both allows different research to simulate environmental time series for simulation studies in a consistent way, it also offers several methods of simulating time series, and so allows researchers to check if simulation results are sensitive to the method of simulating study data. Tools for consistency across studies have helped in the progress of environmental epidemiology research; for example, the public availability of the NMMAPs dataset--which included air pollution, weather, and mortality counts in over 100 US communities for 14 years-- allowed researchers around the world to explore hypotheses and investigate health effects of air pollution and temperature using the same data, and so differences in results could be traced to methodological differences rather than random differences in environmental datasets. [We could add some on the usefulness of (1) reproducible studies and (2) open source software for scientific studies here.]

Further, developing the code to simulate environmental time series in an appropriate way can be time consuming. While researchers often need to customize certain parts of a simulation, to test a particular research question, often most of the simulation can be performed in a way that is generalizable across most other environmental time series studies. Therefore, by providing simulation tools with reasonable defaults, as well as room to customize at each step of the simulation, this package can make environmental epidemiolgoy simulation studies much more efficient and also make simulation a reasonable alternative for determining expected power when planning and proposing future studies.

# Examples of studies that simulated environmental time series

## Mortality displacement

\citet{Armstrong2014} used a simulation study to investigate whether evidence of short-term mortality displacement following heat waves was a result of model artifacts. To do this, they aimed to simulate a time series of heat wave (binary) effects with heat wave effects, none of which were caused by short-term mortality displacement. They then showed that, when they fit the model from the previous study to this simulated dataset, the model generated results suggestive of short-term mortality displacement, and less displacement with more severe heat waves than less severe heat waves, which were similar to the findings of the study they were assessing. Overall, this analysis suggested that the earlier study's results might have been an artifact of the modeling approach they used, and so their results could be consistent with a hypothesis of no mortality displacement.

To do this simulation, which required simulating a time series with a heat wave effect on mortality, but no short-term displacement to explain some of that excess mortality, \citet{Armstrong2014} started with a real dataset of daily temperature and mortality counts. For this, they used the NMMAPS dataset (1987--2000) from Atlanta, GA, which was one of the datasets used in the original study. They fit a model of expected log mortality count for this real data regressed on daily (lag 0) temperature and a smooth function of time (to capture seasonal and long-term variation). They then predicted from this model for the days in their simulation, to generate a smoothed estimate of expected daily mortality on each day (note: this daily expected count is a function of observed temperature over the study period; it sounds like they used observed data directly for the exposure time series). They then drew simulated daily mortality counts from a Poisson distribution with $\lambda$ equal to this daily expected mean mortality for each day in the simulation period. [Code they used for this simulation analysis is available as an online supplement.]

## Comparing case-crossover and time series

## Investigating confounding

## Exploring spatial misalignment between exposure and outcome measurements

@@

### 2016-06-24 writing/SimulationPaper/SimulationPaper.Rmd
Work on paper drafts

@@

@@

@@

@@

>>>>

#### Mortality displacement

Some immediate fatalaties during natural disasters might result from very frail people dying a few days earlier than under the counterfactual of no disaster. For example, [evidence on mortality displacement following heat waves, other disasters]. While these deaths still represent excess premature mortality related to the disaster, they represent a much shorter loss of life-time than premature deaths that were advanced by months or years. Further, in model designs that use times near to the disaster as controls, including time series and case-crossover designs, displaced mortality could cause the estimated baseline mortality during control periods to be artifically low, so biasing the acute disaster effect estimate to be too high [we might want to do a figure that sketches this out].

#### Delayed effects

Disasters could elevate mortality risk in the week or weeks after the acute exposure. For example, [Florida study of mortality in the 2004 hurricane season]. If a study uses controls nearby in time to the disaster, here there would be a bias to underestimate acute disaster effects (the opposite direction of bias than in the case of mortality displacement effects).

A higher-than-average mortality count in the days around a disaster could also be related to autocorrelation in the exposure variable, particularly when a "disaster" is identified as a string of days when a continous exposure outcome is above a certain threshold. For example, @Armstrong2014 note that mortality is expected to be elevated on days around a severe heat wave because temperatures tend to be higher than normal on days just before and after a severe heat wave, and mortality risk is positively associated with temperature across the range of summer temperatures, whether a day meets the threshold for a heat wave or not. A similar pattern could exist for "smokewave" days, used in some studies to explore health risks related to forest fires [Coco's Epidemiology paper].

@@

We generated simulated time series of disaster exposure and mortality outcomes using the `eesim` R package.

We simulated several environmental time series, based on different scenarios of (1) the size of the acute effect of the disaster on mortality risk; (2) the size of the delayed or mortality displacement effect; and (3) the length of either the delayed or mortality displacement effect. We based the range of sizes used for each of these values on reasonable values from studies of the health impacts of heat waves and other natural disasters.

@@

For each simulated time series dataset, we fit three models to estimate acute disaster effects: (1) a time series model; (2) a case-crossover model; and (3) a cross-year model. This last model compares mortality counts during the disaster to a set of control periods within the same city. However, unlike the case-crossover model, it uses control periods from the same month and days of the week as the case in neighboring years, rather than from the same days of the week and month as the case in the same year as the case. This model therefore controls for seasonality by design, but requires addition control to account for year-to-year trends in average mortality.

As a sensitivity analysis, we re-ran our main analysis using environmental time series datasets simulated using some other common simulation methods. We tested results using [other simulation methods].

@@

#### Performance of three models

#### Sensitivity analysis

@@

#### Implications for epidemiological studies of natural disasters

#### More complex models to model delayed or mortality displacement effects

Methods exist to model effect sizes over a period that includes the disaster and the following weeks. [More about dlnm.]

Why researchers might want to model acute effects without using these more complex models:

- Low power (especially with rarer health outcomes, may run into very low power when modeling the more complex models required to assess lags)
- Interest might be in immediate impacts (e.g., what should city planners be prepared for during a disaster, in terms of health care needs, medical examiners?) rather than in estimating the total impacts (e.g., what is the total attributable impact?)
- May not have a good idea of the time range of delayed or mortality displacement effects. If it's longer than the model includes, what are the impacts? Very long time ranges (e.g., month or more) could start to interfere with seasonal patterns, right?

#### Other advantages / disadvantages of each model type

For single disasters, there is the chance for some bias in comparing to controls in the same year if the event occurs during the yearly dip or peak in mortality trends. For example, [Jennifer's ozone and Olympics study]. A model that uses control periods from the same time of the year (i.e., same week) of other years might prevent this bias.

#### Delayed / mortality displacement effects that extend past a year

For very severe disasters, there is some evidence that delayed or mortality displacement effects might span into the following year. [French and Russian heat wave examples, hurricane examples]. In this case, even a study design that uses controls from other years could be biased by how these effects influence control periods.
